Causal inference and co-evolutionary contingencies using dynamic phylogenetic models

### 2022-07-14 manuscript.qmd
working through limits of PGLMMs vis-a-vis state-dependence

## Introduction

@@

## Co-evolution as correlated drift: the evolutionary regression coefficient

@@

The confounder denoted $U$ is unobserved, meaning that we can't statistically adjust for it. However, note that for any time $t$, $X_t$ and $Y_t$ are independent of $U_0$ conditional on $\{ X_{t-1},X_{t-1}$. In other words, the present is conditionally independent of $U_0$ (or "d-seperated") given the past. But in most comparative studies, we have not observed the past, only the present. This is where phylogeny can help us, because more closely related species or populations have more similar confounding factors (e.g., shared genetic architecture, environmental similarity), on average, than distantly related ones. Assuming that we have an appropriate model of evolution, then the patristic distance between species (i.e., time since divergence from a common ancestor) allows us to probabalistically reconstruct unobserved past values of $X$ and $Y$, closing the backdoor path $X \leftarrow U \rightarrow Y$ (cite Rethinking 2nd edition). It should be noted that such a scenario is sometimes called "phylogenetic confounding," but...[DOESN"T WORK IF U IS STILL PRESENT AT THE TIPS. HOW CAN WE TELL IF ITS PHYLOGENETIC CONFOUNDING VS A CAUSAL RELATIONSHIP? IN THE FORMER, THE COVARIANCE BETWEEN X AND Y WILL DECREASE OVER TIME. IN THE LATTER IT IS CONSTANT/STABLE].

@@

Additionally, even if there was zero structural phylogenetic confounding (Figure 1A), it would still be a good idea in any finite sample to adjust for phylogeny because, as a source of variation in $Y$, phlyogeny is a "good control" in that it helps us isolate variation in $Y$ that is due to possible causal effects of $X$. Marginalizing over phylogenetic position can also give us improved estimates of the average effects when clades are unequally sampled.

Now that we have a sense for the assumed causal model underlying most phylogenetic comparative analyses, and motivation for why we statistically adjust for phylogeny, let's review some popular methods for estimating the evolutionary regression coefficient and introduce some variations on the causal model in Figure 1.

### Phylogenetic Generalized Least Squares (PGLS)

@@

$$ W_{T= t} - W_{T = 0} \sim\mathcal{N}(0,\sigma t) $$ Therefore, the variance for a given species or population $j$ is proportional to the total branch length of the tree leading from root to tip, and the covariance between $j$ and any other species or population $k$ is proportional to their amount of shared history. For example, imagine a phylogeny with only $J$ = 3 tip species. The last common ancestor of all species is dated to T = 1,000,000 years ago, and the last common ancestor of 2 and 3 is dated to 0.5T = 500,000 years ago. The implied phylogenetic variance-covariance matrix ($\boldsymbol{V}$) under a Brownian motion model is thus:

@@

Empirically, the actual degree of covariance between related species or populations may be less than expected under a pure Brownian motion model. For example, when evolution is rapid (or more mundanely, when measurement error on trait data is substantial) the "phylogenetic signal" will be weak such that even closely related taxa may have dramatically different trait values [@blomberg2003testing; @kamilar2013phylogenetic]. PGLS can estimate the strength of phylogenetic signal from comparative data, multiplying the off-diagonals of the variance-covariance matrix by $\lambda \in [0,1]$ [@pagel1999inferring]. For example:

@@

Brownian motion is the simplest and most widely used model, and often interpreted as "drift" across the phylogenetic tree. Other popular choices include Ornstein-Uhlenbeck, where change is due to both random drift and "stabilizing selection" around some optimal trait value, as well as Early Burst models of adaptive radiation where the rate of change ($\sigma$) decreases over time. NOTE ON OU COVARIANCE: An OU model where the response is undergoing adaptive evolution and the predictor is drifting also works. But if the predictor is under selection, things get messed up. In general, $\boldsymbol{V}$ is constructed by inputting a matrix of phylogenetic distances to some covariance function $C(x,x')$ such that $V_{[j,k]} = C(j,k)$.

Although PGLS is widely used in comparative biology, it has several key limitations. These include the assumption of Gaussian errors, proneness to overfitting due to a lack of parameter regularization, and failure to accommodate many common data structures such as repeated measures of species trait values, missing data, and measurement error. In the next section we review more flexible regression approaches, as well as an important alternative method of estimating evolutionary associations.

### Phylogenetic Generalized Multilevel Models

Why would we care about the phylogenetic correlation? Because this is an indicator that the traits co-evolve across the phylogenetic tree rather than just at the tips.

Phylogenetic generalized multilevel (or mixed-effects) models (PGLMMs) extend the study of trait coevolution by leveraging analytic strategies developed in quantitative genetics to accommodate more complex, multivariate model designs with non-Gaussian outcomes [@lynch1991methods; @hadfield2010general; @ives2010phylogenetic; @ives2011generalized]. Rather than explicitly scaling the off-diagonals of $\boldsymbol{V}_s$ with some parameter $\lambda$, phylogenetic signal is quantified as the ratio of the variance of phylogenetic random effects $\nu$ (also called 'mixed' or 'varying' effects) to total trait variance. For example, an observation of a Gaussian trait $Y$ from a species or population $j$ and predictor $X$ is modeled as:

@@

In this model, $\beta_{x(\textrm{PGLMM})}$ has the same interpretation as $\beta_{x(\textrm{PGLS})}$. If the response was non-Gaussian, $\beta_{x(\textrm{PGLS})}$ has the same interpretation *on the latent scale*. Conversion of the phylogenetic regression coefficient $\beta_{x(\textrm{PGLMM})}$ to the observation scale requires inverting the link function, and will be sensitive to other model parameters.

### Mutli-response PGLMMs: contrasting macro and microevolution

@@

$$ \rho(X,Y) = \frac{\sigma(X,Y)}{\sigma(x)\sigma(Y)} $$
$$ b = \rho(X,Y)\frac{\sigma(Y)}{\sigma(X)} $$

@@

![Figure 1](figures/phylo_DAG.png)
### Phylogenetic Path Analysis

Independent contrasts, PGLS, PGLMMs, PPA

Discrete data can be modeled in the GLMMM framework if we conceptualize discrete phenotypes (binary, ordinal, categorical) as manifestations of a latent continuous trait, as in the quantitative genetic threshold model [@wright1934; @felsenstein2005using].

## Dynamic state-space models of co-evolution

Pagel's discrete, Hidden Markov Models, Butler & King OU

can be identified statistically as non-zero between-trait phylogenetic correlations----or any other measure of mutual information over evolutionary time.

@@

>>>>

## 1. Introduction

@@

## 2. Co-evolution as correlated drift: the state-independent evolutionary regression coefficient

@@

The confounder denoted $U$ is unobserved, meaning that we can't statistically adjust for it. However, note that for any time $t$, $X_t$ and $Y_t$ are independent of $U_0$ conditional on $\{ X_{t-1},X_{t-1} \}$. In other words, the present is conditionally independent of $U_0$ ("d-seperated") given the past. But in most comparative studies, we have not observed the past, only the present. This is where phylogeny can help us, because more closely related species or populations have more similar confounding factors (e.g., shared genetic architecture, environmental similarity), on average, than distantly related ones. Assuming that we have an appropriate model of evolution, then the patristic distance between species (i.e., time since divergence from a common ancestor) allows us to probabalistically reconstruct unobserved past values of $X$ and $Y$, closing the backdoor path $X \leftarrow U \rightarrow Y$ (cite Rethinking 2nd edition). It should be noted that such a scenario is sometimes called "phylogenetic confounding," but...[DOESN"T WORK IF U IS STILL PRESENT AT THE TIPS. HOW CAN WE TELL IF ITS PHYLOGENETIC CONFOUNDING VS A CAUSAL RELATIONSHIP? IN THE FORMER, THE COVARIANCE BETWEEN X AND Y WILL DECREASE OVER TIME. IN THE LATTER IT IS CONSTANT/STABLE].

@@

Additionally, even if there was zero structural phylogenetic confounding (Figure 1A), it would still be a good idea in any finite sample to adjust for phylogeny because, as a source of variation in $Y$, phylogeny is a "neutral-good control" in that it reduces variation in $Y$, thus improving the precision of our estimates for the causal effect of $X$ [@cinelli2021crash]. Marginalizing over phylogenetic position can also give us improved estimates of the average effects when clades are unequally sampled.

Now that we have a sense for the assumed causal model underlying most phylogenetic comparative analyses, and motivation for why we statistically adjust for phylogeny, let's review some popular methods for estimating the evolutionary regression coefficient, and introduce some variations on the causal model in Figure 1.

### 2.1 Phylogenetic Generalized Least Squares (PGLS)

@@

$$ W_{T= t} - W_{T = 0} \sim\mathcal{N}(0,\sigma t) $$ Therefore, the variance for a given species or population $j$ is proportional to the total branch length of the tree leading from root to tip, and the covariance between $j$ and any other species or population $k$ is proportional to their amount of shared history. Note that this is a continuous-time analog of the discrete time model presented in Figure 1. For example, imagine a phylogeny with only $J$ = 3 tip species. The last common ancestor of all species is dated to T = 1,000,000 years ago, and the last common ancestor of 2 and 3 is dated to 0.5T = 500,000 years ago. The implied phylogenetic variance-covariance matrix ($\boldsymbol{V}$) under a Brownian motion model is thus:

@@

Empirically, the actual degree of covariance between related species or populations may be less than expected under a pure Brownian motion model. For example, when evolution is rapid (or more mundanely, when measurement error is substantial) the "phylogenetic signal" will be weak such that even closely related taxa may have dramatically different trait values [@blomberg2003testing; @kamilar2013phylogenetic]. PGLS can estimate the strength of phylogenetic signal from comparative data, multiplying the off-diagonals of the variance-covariance matrix by $\lambda \in [0,1]$ [@pagel1999inferring]. For example:

@@

Brownian motion is the simplest and most widely used model, and often interpreted as "drift" across the phylogenetic tree. Given the assumptions laid out in the previous section, the Brownian motion covariance function allows us to appropriately adjust for phylogenetic confounding and non-independence. But if traits do not evolve according to a Brownian motion process, we need a different way to construct $V$. In general, $\boldsymbol{V}$ is constructed by inputting a matrix of patristic distances to some covariance function $C(x,x')$ such that $V_{[j,k]} = C(j,k)$. For example, Early Burst models of adaptive radiation allow the rate of change ($\sigma$) to decrease over time. Another popular choice is the Ornstein-Uhlenbeck (OU) covariance function, which follow from a model of trait evolution where change is due to both stochastic "drift" and deterministic "selection." As a result, the relationship between patristic distance and phylogenetic covariance becomes non-linear. Note however that use of an OU covariance function in the PGLS framework still assumes zero selection on the predictors.

Although PGLS is widely used in comparative biology, it has several key limitations. These include the assumption of Gaussian errors, proneness to overfitting due to a lack of parameter regularization, and failure to accommodate many common data features such as repeated measures of species trait values, missing data, and measurement error. In the next section we review more flexible regression approaches, as well as an important alternative method of estimating evolutionary associations that distinguishes between macro and micro evolutionary correlations.

### 2.2 Phylogenetic Generalized Multilevel Models (PGLMMs)

Phylogenetic generalized multilevel (or mixed-effects) models (PGLMMs) extend the study of trait coevolution by leveraging analytic strategies developed in quantitative genetics to accommodate more complex model designs with non-Gaussian outcomes [@lynch1991methods; @hadfield2010general; @ives2010phylogenetic; @ives2011generalized]. Rather than explicitly scaling the off-diagonals of $\boldsymbol{V}_s$ with some parameter $\lambda$, phylogenetic signal is quantified as the ratio of the variance of phylogenetic random effects $\nu$ (also called 'mixed' or 'varying' effects) to total trait variance. For example, an observation of a Gaussian trait $Y$ from a species or population $j$ and predictor $X$ is modeled as:

@@

In this model, $\beta_{x(\textrm{PGLMM})}$ has the same interpretation as $\beta_{x(\textrm{PGLS})}$. If the response was non-Gaussian, $\beta_{x(\textrm{PGLS})}$ has the same interpretation *on the latent scale*. Conversion of the phylogenetic regression coefficient $\beta_{x(\textrm{PGLMM})}$ to the observation scale requires inverting the link function, and will be sensitive to other model parameters. Modelling discrete data in the PGLMMM framework can be biologically motivated if we conceptualize discrete phenotypes (binary, ordinal, categorical) as manifestations of a latent continuous trait, as in the quantitative genetic threshold model [@wright1934analysis; @felsenstein2005using].

In addition to accommodating non-Gaussian traits, PGLMMs can readily account for measurement error, within-species trait variation, phylogenetic uncertainty, and missing data [@harmon2005effect; @garamszegi2014uncertainties; @jaeggi2020variable]. Under the PGLMM umbrella, authors have also extended evolutionary models of trait change to include variable rates of drift and directional change across the phylogenetic tree [@revell2021variable; @pagel2022general].

### 2.3 Mutlivariate PGLMMs: partitioning macro and microevolution

@@

$$ \rho(X,Y) = \frac{\sigma(X,Y)}{\sigma(x)\sigma(Y)} $$
$$ b = \rho(X,Y)\frac{\sigma(Y)}{\sigma(X)} $$
The variance partioning approach can be useful, but to study adaptive co-evolution we will need to move beyond the GLMM world, utilizing state-space models.

@@

![Figure 1](figures/phylo_DAG.png)
### 2.4 Phylogenetic Path Analysis

## 3. Dynamic state-space models of co-evolution

### 3.1 Ornstein-Uhlenbeck and co

### 3.2 Discrete, HMM

Pagel's classic limited to two binary traits, diveristree and HMM add flexibility to more than two states and variable transition rates throughout the tree, but remain limited in the discrete setting. Given infinite data, discrete transition matrix is good, but in finite samples it is problematic.... Unfortunately though this method is fundamentally limited to discrete traits, sometimes forcing researchers to dichotomize continuous variables (e.g. environmental harshness [@cornwallis2017cooperation]), which can lead to loss of power and inconsistent results [@dawson2012dichotomizing; @royston2006dichotomizing]. Because of this limitation to binary traits this method has not been applied to topics typically involving continuous traits (e.g. morphology, life history, brain size, etc.).

### 3.3 GDPM: how our method combines features of OU and Discrete, and PGLMMs

@@

IDEA: In simulation study, compare statistical power (PP > 0.9, or whatever) vs phylogenetic signal.

## 4. Case study: Primate life history and niche complexity
